---
title: "Programming in R"
subtitle: "Part 1: Fundamentals"
institute: "DS Collab"
author: "Ethan Marzban"
format: html
# title-block-banner: "#4144a6"
title-block-banner: logo.png
css: styles.scss
---

## Prerequisites

- Please make sure you have downloaded both `R` and `RStudio`, as outlined in the previous lab.

## What Is Programming?

Computers are an incredibly useful tool in a Data Scientist's arsenal. They are, however, also incredibly complex and can be difficult to communicate with. As such, **programming languages** are used to help us communicate with computers and provide them with instructions on what we want them to do. There are several different programming languages: the two most often used in Data Science circles are **`R`** and **`Python`**, though other popular languages include **`Julia`**, **`MatLab`**, and **`C+`**. 

Computer programs can be written in a variety of different environments and **editors**; for example, Python can be written in `Jupyter` Notebooks, VS Code, and several others. `R` is most commonly run in **`RStudio`**, which can be downloaded for free (along with the programming language `R`). Please see the previous lab for further instructions on how to download `R` and `RStudio`.

## Programming Quick Start

There is a reason we use the word "language" to describe programming languages- that is because they function quite like a human language. This means, among other things, that they each have their own **syntax** (i.e. set of grammar rules). 

Programs are made up of **expressions**, like `2 + 2`. We **evaluate** expressions by **running** (or **executing**) them in a programming language. Expressions are like the sentences of programming- they contain complex pieces of information that are conveyed between the user and the computer.

Much like sentences in other languages, expressions must obey a rigid syntax. For example, when we want to perform addition in `R` we must use the `+` symbol; we can't, for example, say `2 plus 2` and expect `R` to know what to do.

Speaking of addition, one of the easiest ways to start using `R` is to use it as a calculator! `R`, much like many other programming languages, obeys the standard **order of operations** when evaluating expressions:

- **P**arentheses
- **E**xponents
- **M**ultiplication
- **D**ivision
- **A**ddition
- **S**ubtraction

Here is a list of mathematical operators and their corresponding `R` syntax:

| **Operation** | **`R` Operator** | **Example** | **Result** |
|:--------|:--------|:---------|:--------|
| Addition | `+` | `2 + 2` | `4` |
| Subtraction | `-` | `2 - 2` | `0` |
| Multiplication | `*` | `2 * 2` | `4` |
| Division | `/` | `2 / 2` | `1` |
| Exponentiation | `^` | `2 ^ 2` | `4` |


:::{.callout}
## **Exercise 1**

Evaluate $\displaystyle \left(\frac{2 + 3^{3/2}}{4}\right)^{2}$ using `R`.
:::

## Variable Assignment

Let's talk a bit about **variables**. Just like in math, variables in a programming language function as a sort of placeholder for a particular piece of information (be it a function, value, etc.) The act of storing information in a variable is called **assignment**, and in `R` variable assignment is performed using the `<-` symbol.

```{r, eval = F}
<variable name> <- <what you want to associate with the variable>
```

For example, after running
```{r}
x <- 2
```
the quantity `x` will always be synonymous with the quantity `2`, and running `x + 2` will return a value of `4` (as 2 + 2 = 4).

`R` affords a lot of flexibility when it comes to variable names- that is, we can pick almost anything we want to be a variable name! There are, however, some exceptions:

- Variable names cannot start with a number
- Variable names cannot include a space

:::{.callout-tip}
It is a good programming practice to give your variables names that are descriptive, but not overly long.
:::

If we want to view the value stored in a variable, we have two options: we could simply type the name of the variable, and run the cell:


```{r}
x
```

or we could pass the variable name into a call to the `print()` function (we'll talk more about functions in a future workshop):

```{r}
print(x)
```

:::{.callout-caution}
Variable names in `R` are case sensitive.
:::

For example, if you run
```{r}
my_variable <- 25
```
`R` will *not* know what to do with the variable `My_variable`!
```{r, error = T}
My_variable
```


:::{.callout-tip}
The errors `R` outputs often contain useful information- be sure to read them!
:::

Sometimes it will be necessary to **update** or **re-assign** a new value to an existing variable. For example, let’s examine the structure of the following code:
```{r}
x = 2
x = x + 3
```


What do you think running `x` will return? If you said `5`, you’d be correct! The key point of this is:

:::{.callout-note}
## Important
When performing variable assignment, `R` reads from right to left.
:::

So, in code example above, `R` first executed `x + 3` (which is equivalent to `2 + 3`; i.e. `5`), and then re-assigned `x` the value `5`.


:::{.callout}
## **Exercise 2**

a) Define a variable called `num_sections`, and assign it the value of `3`.
b) Define another variable called `section_capacity`, and assign it the value of `25`.
c) Update the value of `num_sections` to be one higher than it originally was.
d) Compute the product of `num_sections` and `section_capacity`.
:::


## Comments

Code scripts can get long and complicated, pretty quickly. As such, it is often a good idea to add **comments** to your code. Comments are pretty much exactly what they sound like- they are short words or phrases that do *not* get executed, but can help the reader understand the code better.

In `R`, comments are created using the hashtag (`#`). For example:
```{r}
x <- 1      # define x
x <- x + 1  # increment x by 1
```
Again, the phrases `define x` and `increment x by 1` are never executed by `R`; they exist solely to help the reader understand what each line of code is doing.

## Functions

Functions in `R` behave much like functions in mathematics. 

Given a function `f()` that has been defined in `R` (we will talk about how to define functions in a future workshop), we **call** the function on **inputs** (aka **arguments**) `arg1`, `arg2`, ... using the syntax
```{r, eval = F}
f(arg1, arg2, ...)
```

For example, as we saw previously, the `print()` function can be called on an argument to simply *print* the output:
```{r}
print("hello")
```

## Help Files

`R` is an **open-source** software, meaning there are continually contributions being made. This means that there is no way to know *everything* about `R`! Rather, we must rely on the extensive **documentation** that is provided with most `R` packages and functions. 

To access the **help file** on a particular function, you can use the syntax `?<function name>`. For example, `?plot` pulls up the help file for the `plot()` function in `R`.

:::{.callout}
## **Exercise 3**

Look up the help file for the `sin()` function.
:::

Sometimes, we may want to look up the help file for function whose name we don't know exactly. To look up help files containing a word, we can use `??<name>`. 

:::{.callout}
## **Exercise 4**

Run `??exponent`, and find out how to perform exponentiation in `R`. Use this to compute $e^{4.2}$.
:::


## Packages

In `R`, **packages** can be thought of as bundles containing several functions and/or structures. By default, `R` loads (i.e. includes) what is known as the **base `R`** package, which comes equipped with several useful functions. 

We can access help files for particular packages using the same syntax as we used to look up the help files for functions: `?<package name>`. 

:::{.callout}
## **Exercise 5**

Look up the help file for the `base` package. Then, run code to display a list of *all* functions included in the `base` package.
:::

:::{.callout-tip}
Help files for packages are often referred to as **vignettes** by programmers who routinely use `R`.
:::

With a few exceptions, we always need to **install** a package before using it for the first time. The syntax we use to install a package with the name `package_name` is
```{r, eval = F}
install.packages(package_name)
```

:::{.callout-caution}
You only need to install a package **once**.
:::

After installing a package, we need to **load** or **import** it.  The syntax we use to load a package with the name `package_name` is
```{r, eval = F}
library(package_name)
```